\thornado-hydro: Generalizing discontinuous Galerkin methods for a nuclear equation of state for supernova hydrodynamics

### 2019-05-05 paper/ms.tex
Working through final revisions. Final push.

\title{\thornado-hydro: Generalizing discontinuous galerkin methods for a nuclear
equation of state for supernova hydrodynamics}
\author{Brandon Barker}
\affiliation{Department of Physics and Astronomy, University of Tennessee Knoxville, TN 37996}

@@

\begin{abstract}
A problem of high importance in computational astrophysics is obtaining accurate
solutions to the Euler equations of hydrodynamics. We are interested in solving
the Euler equations in the context of core collapse supernovae.
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado)
is being developed for CCSN simulations and related problems utilizing a spatial discretization
based on the discontinuous Galerkin (DG) method.
The Euler equations form a hyperbolic set of partial differential equations.
In the quasi-linear form, the system can be represented
as a set of independent advection equations that can be limited separately.
This use of characteristic variables also increases the efficiency of the slope
limiting process. However, the addition of the nuclear equation of state equation
to the Euler equations makes the decomposition nontrivial.
We introduce the framework for the characteristic decomposition of the Euler equations
with the inclusion of the nuclear EOS terms and
present results from some initial tests. The results confirm that performing limiting
on the characteristic variables provides better numerical solutions and is
suited to applications in CCSN simulations.
\end{abstract}

@@

Discontinuous Galerkin (DG) methods have been shown by others
\citep[see e.g.,][]{radice:2011, schaal:2015a, zanotti:2015, dumbser:2018} to have a
high potential for apllications to astrophysical problems. They are a desireable
choice for modelling fluid flows in CCSNe. These methods combine elements of
spectral and finite volume methods, and achieve high-order accuracy
on a compact stencil.

The {\bf t}oolkit for {\bf h}igh-{\bf or}der {\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics (\thornado) is being developed to simulate neutrino-radiation hydrodynamics in CCSNe and related applications in nuclear astrophysics.
The spatial discretization of solvers for hyperbolic partial differential equations in \thornado\, is based on the DG method.
Whether the high-order approach will improve accuracy and efficiency of CCSN models remains to be demonstrated.
In this paper we provide an initial description and encouraging demonstration
of the DG method implemented in \thornado\, to solve the non-relativistic Euler equations.
We focus on basic one-dimensional tests focusing on the implementation of limiters in \thornado.

The DG slope limiting techniques implemented in \thornado\, are similar to those
used by \cite{schaal:2015a}, but is supplemented with a troubled-cell
indicator \cite{fu:2017} to prevent excessive limiting (e.g., around smooth extrema).
It has been shown \citep[see e.g.,][]{schaal:2015a, cockburn:1998, cockburn:1989}
that limiting on the chracteristic variables, which represent advected quantities,
shows an improvement over limiting on the conserved variables.
In this paper, we develop a chracteristic limiter for use with a general tabulated
nuclear matter EOS in \thornado.

This paper will be organized as follows: in Section~\ref{sec:eulereq} we briefly
describe the Euler equations of gas dynamics. Section~\ref{sec:DG} describes our
numerical implementation of the DG method in the one dimensional (1D) case.
Section~\ref{sec:eigen} presents the eigenvectors used in the eigen-decomposition
of the Euler equations, and Section~\ref{sec:results}
shows the results of preliminary 1D hydrodynamic simulations.

\section{Euler Equations of Gas Dynamics in Cartesian Coordinates}
\label{sec:eulereq}
The non-relativistic Euler equations of gas dynamics
\citep[see, e.g.,][for details]{leveque:2002} in Cartesian coordinates
in the absence of sources with a nuclear matter EOS are given by
the equations of conservation of mass,
\beq
  \partial_{t} \rho + \divergence{} (\rho\,  \mathbf{v}) = 0
  \label{eq:massConservation}
\eeq
conservation of momentum,
\beq
%	\partial_{t}(\rho \mathbf{v}) + \divergence{}(\rho u^2 + p) = 0
  \partial_{t}(\rho\,  \mathbf{v}) + \divergence{}(\rho\,  \mathbf{v} \otimes \mathbf{v} + P\, \mathbf{I}) = 0
  \label{eq:momentumConservation}
\eeq
conservation of energy,
\beq
  \partial_{t} E + \divergence{}\left[(E+P)\, \mathbf{v}\right] = 0
  \label{eq:energyConservation}
\eeq
and conservation of \textcolor{red}{electron density}
\beq
  \partial_{t} D_{e} + \divergence{}(D_{e}\, \mathbf{v}) = 0
  \label{eq:electronConservation}
\eeq
where $\rho$ represents mass density, $\mathbf{v}$ the fluid velocity vector,
$P$ the fluid pressure, $D_{e}=\rho y_e$ where $y_e$ is the electron fraction,
$E=\epsilon \rho +\frac{1}{2}\rho v^2$ the total energy (internal plus kinetic),
$\epsilon$ is the specific internal energy, and $\mathbf{I}$ is the identity tensor.
\eqref{eq:massConservation}-\eqref{eq:electronConservation}. The inclusion of
Equation \eqref{eq:electronConservation} is because we require a nuclear EOS.
These equations are closed by a tabulated EOS where the pressure is given by a function of
density, temperature $T$ and the electron fraction: $P = P(\rho, T, y_e)$.
We may rewrite Equations \eqref{eq:massConservation}-\eqref{eq:electronConservation}
in a more convenient way:

\beq
  \partial_{t}\mathbf{U} + \partial_{x}\mathbf{F}(\mathbf{U}) = 0,
  \label{eq:conservation}
\eeq
where $\mathbf{U} =(\rho,\rho \mathbf{v},E, D)^{T}$ is the vector of conserved quantities
and $\mathbf{F}(\mathbf{U})=(\rho \mathbf{v},\rho u^{2}+p,(E+P)\mathbf{v}, D\mathbf{v})^{T}$
is the flux vector.

@@

\subsection{The Discontinuous Galerkin Methods}
In our solver we have chosen the discontinuous Galerkin (DG) method
for our spatial discretization. In this section we will briefly discuss our
implementation of the DG method, introducing notation and concepts.
For simplicity, we will focus on the one dimensional (1D) case. Recall that
we seek solutions to the Euler equations of hydrodynamics, constituting a
hyperbolic conservation law of the form

\beq
  \partial_{t} \mathbf{U} + \partial_{x} \mathbf{F}(\mathbf{U}) = 0
  \label{eq:ode}
\eeq
Where $\mathbf{U}$ is the evolved state vector and $\mathbf{F}(\mathbf{U})$ is
the flux. In order to solve Equation \eqref{eq:ode} numerically, we divide the computational
domain $D\subset \mathbb{R}$ into a disjoint union $\mathcal{T}$ of open elements
$\bK$ such that $D = \cup_{\bK \in \cT}\bK$. Each element $\bK$ is a box in the
coordinates
\beq
  \bK \in (\xL,x_{R}),
\eeq
We let the approximation space $\mathbb{V}^{k}$ for the DG method
be constructed from the tensor product of one-dimensional
polynomials of degree $k$. Note that functions in $\mathbb{V}^{k}$
can be discontinuous across element interfaces. The DG problem is then to find
$\mathbf{U}_h \in \mathbb{V}^{k}$ which approximates $\mathbf{U}$ in Equation
\eqref{eq:ode}, such that $\forall$ $\phi \in \mathbb{V}^{k}$ and $\bK \in \cT$

@@

In each element $K$, we use a nodal representation of the conserved variables
$\mathbf{U}$:

\beq
\begin{split}
  \mathbf{U}(x,t) \approx\, & \mathbf{U}_h (x,t) =
  \sum_{i=1}^{N}\vect{U}_{i}(t)\,\ell_{i}(x),
  \quad\text{where}\quad \\
  & \ell_{i}(\eta)=
  \prod_{\substack{j=1\\j\ne i}}^{N}\f{\eta-\eta_{j}}{\eta_{i}-\eta_{j}}
  \label{eq:conservedNodalExpansion}
\end{split}
\eeq
are Lagrange polynomials defined on $I = \{ \eta : \eta \in (-0.5,0.5) \}$,
and are constructed to interpolate the node set
$S_{N}=\{\eta_{i}\}_{i=1}^{N}\subset I$. The spatial coordinate $x$ and the
reference coordinate $\eta$ are related by the
mapping $x(\eta)=\xL+(0.5+\eta)\,\Delta x$.
Then, for any $\eta_{j}\in S_{N}$, $\ell_{i}(\eta_{j})=\delta_{ij}$,
so that $\vect{U}_{h}(x(\eta_{j}),t)=\vect{U}_{j}(t)$.
First we define the $M$-point quadrature $Q_{M}:C^{0}(I)\to\mathbb{R}$
with abscissas $\hat{S}_{M}=\{\eta_{q}\}_{q=1}^{M}$ and weights
$\{w_{q}\}_{q=1}^{M}$, normalized such that $\sum_{q=1}^{M}w_{q}=1$ in order to
evaluate the integrals in Eq.~\eqref{eq:semidiscreteDG}.
We use the $M$-point Legendre-Gauss quadrature, which is exact for
polynomials of degree $\le 2M-1$.
Then, if $P_{h}(x)$ is such a polynomial, we have
\beq
  \f{1}{\Delta x}\int_{K}P_{h}(x)\,dx=\int_{I}P_{h}(\eta)\,d\eta=\sum_{q=1}^{M}w_{q}\,P_{h}(\eta_{q}).
\eeq

For the sake of efficiency we let $M=N$ and $S_{N}=\hat{S}_{N}$, which is a
spectral-type nodal collocation DG approximation \citep{bassi:2013}.
Inserting Eq.~\eqref{eq:conservedNodalExpansion} into Eq.~\eqref{eq:semidiscreteDG},
letting $\phi(x)=\ell_{k}(x)$, using the quadratures defined above,
we obtain
\begin{align}
  \pd{}{t}\int_{K}\vect{U}_{h}\,\phi\,dx
  &\approx w_{k}\,\pd{}{t}\vect{U}_{k}\,\Delta x
  \label{eq:timeDerivativeTerm}
\end{align}
for the time derivative piece.
Similarly, the last term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}
becomes
\beq
  \int_{K}\vect{F}(\vect{U}_{h})\,\pderiv{\phi}{x}\,dx
  \approx \sum_{q=1}^{N}w_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q}).
  \label{eq:volumeTerm}
\eeq
Now we may combine Eqs.~\eqref{eq:timeDerivativeTerm}-\eqref{eq:volumeTerm} in
Eq.~\eqref{eq:semidiscreteDG} resulting in the semi-discrete form
\beq
  \deriv{\vect{U}_{k}}{t}
   =-\f{1}{w_{k}\Delta x}
  \Big\{
  \Big[\,
    \widehat{\vect{F}}\ell_{k}\big|_{x_{H}}
     -\widehat{\vect{F}}\ell_{k}\big|_{\xL}
  \,\Big]
   -\sum_{q=1}^{N}w_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  \Big\}.
  \label{eq:semidiscreteDiscretized}
\eeq
We now have a system of ordinary
differential equations (ODEs), which may be evolved in time with an ODE solver.
In Section~\ref{sec:results} we use the third-order strong
stability-preserving Runge-Kutta (SSP-RK3) method \citep{shu:1988}.

\subsection{Slope Limiting}
\label{sec:Limiting}
A common feature of high-order numerical PDEs is unphysical oscillations in the solutions,
particularly near discontinuities.
It is therfore of great interest in the DG algorithm to implement slope limiting
of the polynomial $\mathbf{U}_h$. We use the total variation diminishing (TVD)
slope limiter \citep[see, e.g.,][]{cockburn:1998} in conjunction with the
troubled cell indicator discussed in \citet{fu:2017} to prevent excessive limiting
by flagging elements where limiting is needed. To do this, we need to reduce under- and
overshootings of the higher-order solution at cell boundaries compared to
the cell averages of neighbor cells. Recall from Eq.~\eqref{eq:conservedNodalExpansion}
that in each cell our solution is expressed as

\beq
\mathbf{U}_h (x,t) =
\sum_{l=1}^{N}c_{l}(t)\,P_{l}(x)
\eeq

\noindent where we have chosen to expand in the Legendre polynomials. We compare
the weight $c_{1}$, which is proportional to the first derivative of the
solution in the cell, to the neighboring cell averages by the following \\

\beq
  \widetilde{c}_{1} = \text{minmod}(c_{1},
    \beta_{\TVD}(c_{0}^{+} - c_{0}), \beta_{\TVD}(c_{0} - c_{0}^{-}))
  \label{eq:limiting}
\eeq

\noindent where $\widetilde{c}_{1}$ is the limited weight. The superscripts $-/+$
indicate the cell averages of the neighboring cells to the  immediate left/right
of the interface.
Here, we use the minmod function defined as

@@

Limiting is applied to each component of the vector of conserved quantities
separately and is generally referred to as `Componentwise limiting.`
The parameter $\beta_{\TVD}$ takes values in the closed interval $\left[1,2\right]$
and scales the strength of the limiting. A minimal $\beta_{\TVD}$ corresponds with
the total variation diminishing scheme \textcolor{red}{reference?} but is more
dissipative than the maximal $\beta_{\TVD}$ case, which is more flexible but allows for more oscillations.
Increasing $\beta_{\TVD}$ puts more weight on the neighboring cell averages,
making the minmod function more likely to set $ \widetilde{c}_{1} = c_{1}$.

In order to determine where slope limiting is necessary, we use the
troubled cell indicator (TCI) \citep{fu:2017} to prevent excessive limiting:\footnote{Would this notation be altered for 1D?}

@@

\noindent where $G\in\vect{G}\subseteq\vect{U}$ and $\vect{G}$ consists of the
first and final elements of $\vect{U}$. Here, the sum in the numerator and the
max in the denominator are taken over the neighboring elements sharing a
boundary with target element $\bK$, $G_{\bK}$ is the cell average in $\bK$,
$G_{\bK}^{(j)}$ is the cell average
computed by extrapolating the polynomial representation from the neighboring
element $\bK^{(j)}$ into $\bK$, and $G_{\bK^{(j)}}^{(j)}$
is the cell average native to neighbor element $\bK^{(j)}$. An element is flagged
for limiting if, for any $G\in\vect{G}$, $I_{\bK}(G)>C_{\TCI}$, where
$C_{\TCI}$ is a defined threshold.

>>>>

\title{\thornado-hydro: Generalizing discontinuous Galerkin methods for a nuclear
equation of state for supernova hydrodynamics}
\author{Brandon Barker}
\affiliation{Department of Physics and Astronomy, University of Tennessee Knoxville, TN 37996}

@@

\begin{abstract}
A problem of high importance in computational astrophysics is obtaining accurate
solutions to the Euler equations of hydrodynamics. We are interested in solving
the Euler equations in the context of core collapse supernovae.
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado)
is being developed for core-collapse supernova (CCSN) simulations and related problems utilizing a spatial discretization
based on the discontinuous Galerkin (DG) method.
The Euler equations form a hyperbolic set of partial differential equations.
In the quasi-linear form, the system can be represented
as a set of independent advection equations that can be limited separately.
This use of characteristic variables also increases the efficiency of the slope
limiting process for high-order DG methods. However, the addition of a
tabulated nuclear equation of state to the Euler equations makes the decomposition nontrivial.
We introduce the framework for the characteristic decomposition of the Euler equations
with the inclusion of the nuclear EOS terms and
present results from some initial tests using a third-order scheme.
The results confirm that performing limiting
on the characteristic variables provides better numerical solutions and is perhaps
suited to applications in CCSN simulations.
\end{abstract}

@@

Discontinuous Galerkin (DG) methods have been shown by others
\citep[see e.g.,][]{radice:2011, schaal:2015a, zanotti:2015, dumbser:2018} to have a
high potential for applications to astrophysical problems. They are a desireable
choice for modelling fluid flows in CCSNe. These methods combine elements of
spectral and finite volume methods, and achieve high-order accuracy
on a compact stencil.

The {\bf t}oolkit for {\bf h}igh-{\bf or}der {\bf n}eutrino-r{\bf ad}iation hydr{\bf o}dynamics
(\thornado)\footnote{\url{https://github.com/endeve/thornado}}is being developed to simulate neutrino-radiation
hydrodynamics in CCSNe and related applications in nuclear astrophysics.
The spatial discretization of solvers for hyperbolic partial differential equations in \thornado\, is based on the DG method.
Whether the high-order approach will improve accuracy and efficiency of CCSN models remains to be demonstrated.
In this paper we provide an initial description and encouraging demonstration
of the DG method implemented in \thornado\, to solve the non-relativistic Euler equations.
We focus on basic one-dimensional tests focusing on the implementation of limiters in \thornado.

The DG slope limiting techniques implemented in \thornado\, are similar to those
used by \cite{schaal:2015a}, but is supplemented with a troubled-cell
indicator \citep{fu:2017} to prevent excessive limiting (e.g., around smooth extrema).
It has been shown \citep[see e.g.,][]{schaal:2015a, cockburn:1998, cockburn:1989}
that limiting on the chracteristic variables (characteristic limiting),
which represent advected quantities, shows an improvement over limiting
directly on the conserved variables (componentwise limiting).
In this paper, we develop a chracteristic
limiter for use with a general tabulated nuclear matter EOS in \thornado.
In the case of the ideal EOS, the eigenvectors and eigenvalues which diagonalize
the flux Jacobian, which are necessary to transform to the characteristic variables,
are known analytically. All of the thermodynamic derivatives have analytic forms.
The inclusion of a tabulated nuclear EOS complicates matters, as derivatives
must be appriximated by interpolating from the EOS table and the diagonalization
becomes non-trovial.
We provide explicit analytic expressions for the matrices which diagonalize the
flux Jacobian of the Euler equations.

This paper is organized as follows: in Section~\ref{sec:eulereq} we briefly
describe the Euler equations of gas dynamics including a conservation law
appropriate for a nuclear EOS. Section~\ref{sec:DG} describes our
numerical implementation of the DG method in the one dimensional (1D) case.
Section~\ref{sec:eigen} presents the eigenvectors used in the diagonalization
of the flux Jacobian of the Euler equations, and Section~\ref{sec:results}
shows the results of preliminary 1D hydrodynamic Riemann problem similar to that
of Sod \citep{sod:1978}.

\section{Euler Equations of Gas Dynamics in Cartesian Coordinates}
\label{sec:eulereq}
The non-relativistic Euler equations of gas dynamics
\citep[see, e.g.,][for details in the case of an ideal EOS]{leveque:2002}
in the absence of sources with a nuclear matter EOS are given by
the equations of conservation of mass,
\beq
  \partial_{t} \rho + \divergence{} (\rho\,  \mathbf{v}) = 0
  \label{eq:massConservation}
\eeq
conservation of momentum,
\beq
%	\partial_{t}(\rho \mathbf{v}) + \divergence{}(\rho u^2 + p) = 0
  \partial_{t}(\rho\,  \mathbf{v}) + \divergence{}(\rho\,  \mathbf{v} \otimes \mathbf{v} + P\, \mathbf{I}) = 0
  \label{eq:momentumConservation}
\eeq
conservation of energy,
\beq
  \partial_{t} E + \divergence{}\left[(E+P)\, \mathbf{v}\right] = 0
  \label{eq:energyConservation}
\eeq
and conservation of electrons
\beq
  \partial_{t} D_{e} + \divergence{}(D_{e}\, \mathbf{v}) = 0
  \label{eq:electronConservation}
\eeq
where $\rho$ represents mass density, $\mathbf{v}$ the fluid velocity vector,
$P$ the fluid pressure, $D_{e}=\rho y_e$ where $y_e$ is the electron fraction,
$E=\epsilon \rho +\frac{1}{2}\rho v^2$ the total energy (internal plus kinetic),
$\epsilon$ is the specific internal energy, and $\mathbf{I}$ is the identity tensor.
\eqref{eq:massConservation}-\eqref{eq:electronConservation}. The inclusion of
Equation \eqref{eq:electronConservation} is because we require a nuclear EOS.
These equations are closed by a tabulated EOS where the pressure is given by a function of
density, temperature $T$ and the electron fraction: $P = P(\rho, T, y_e)$.
We may rewrite Equations \eqref{eq:massConservation}-\eqref{eq:electronConservation}
in a more convenient way:

\beq
  \partial_{t}\mathbf{U} + \divergence{}\mathbf{F}(\mathbf{U}) = 0,
  \label{eq:conservation}
\eeq
where $\mathbf{U} =(\rho,\rho \mathbf{v},E, D_{e})^{T}$ is the vector of conserved quantities
and $\mathbf{F}(\mathbf{U})=(\rho \mathbf{v},
\rho\, \mathbf{v} \otimes \mathbf{v} + P\, \mathbf{I},(E+P)\mathbf{v}, D_{e}\mathbf{v})^{T}$
is the flux vector.

@@

\subsection{The Discontinuous Galerkin Methods}
In our solver we have chosen the discontinuous Galerkin (DG) method
for our spatial discretization \citep[see e.g.,][]{schaal:2015, zhang:2010a, cockburn:1998, cockburn:1989}.
In this section we will briefly discuss our
implementation of the DG method, introducing notation and concepts.
For simplicity, we will focus on the one dimensional (1D) case. Recall that
we seek solutions to the Euler equations of hydrodynamics, constituting a
hyperbolic conservation law of the form

\beq
  \partial_{t} \mathbf{U} + \partial_{x} \mathbf{F}(\mathbf{U}) = 0
  \label{eq:ode}
\eeq
Where $\mathbf{U}$ is the evolved state vector and $\mathbf{F}(\mathbf{U})$ is
the flux. In order to solve Equation \eqref{eq:ode} numerically, we divide the computational
domain $D\subset \mathbb{R}$ into a disjoint union $\mathcal{T}$ of open elements
$\bK$ such that $D = \cup_{\bK \in \cT}\bK$. Each element $\bK$ is a box in the
coordinates
\beq
  \bK \in (\xL,x_{R}),
\eeq
where $\xL$ and $x_{R}$ are the left and right boundaries of the cell.
We let the approximation space $\mathbb{V}^{k}$ for the DG method
be polynomials of maximal degree $k$. Note that functions in $\mathbb{V}^{k}$
can be discontinuous across element interfaces. The DG problem is then to find
$\mathbf{U}_h \in \mathbb{V}^{k}$ which approximates $\mathbf{U}$ in Equation
\eqref{eq:ode}, such that $\forall$ $\phi \in \mathbb{V}^{k}$ and $\bK \in \cT$

@@

In each element $\bK$, we use a nodal representation of the conserved variables
$\mathbf{U}$:

\begin{equation}
  \vect{U}(x,t)\approx
  \vect{U}_{h}(x,t)=\sum_{i=1}^{N=k+1}\vect{U}_{i}(t)\,\ell_{i}(x),
  \quad\text{where}\quad
  \ell_{i}(\eta)=
  \prod_{\substack{j=1\\j\ne i}}^{N=k+1}\f{\eta-\eta_{j}}{\eta_{i}-\eta_{j}}
  \label{eq:conservedNodalExpansion}
\end{equation}

are Lagrange polynomials defined on $I = \{ \eta : \eta \in (-0.5,0.5) \}$,
and are constructed to interpolate the node set
$S_{N}=\{\eta_{i}\}_{i=1}^{N}\subset I$. The spatial coordinate $x$ and the
reference coordinate $\eta$ are related by the
mapping $x(\eta)=\xL+(0.5+\eta)\,\Delta x$.
Then, for any $\eta_{j}\in S_{N}$, $\ell_{i}(\eta_{j})=\delta_{ij}$,
so that $\vect{U}_{h}(x(\eta_{j}),t)=\vect{U}_{j}(t)$.
We define the $M$-point quadrature $Q_{M}:C^{0}(I)\to\mathbb{R}$
with abscissas $\hat{S}_{M}=\{\eta_{q}\}_{q=1}^{M}$ and weights
$\{w_{q}\}_{q=1}^{M}$, normalized such that $\sum_{q=1}^{M}w_{q}=1$ in order to
evaluate the integrals in Eq.~\eqref{eq:semidiscreteDG}.
We use the $M$-point Legendre-Gauss quadrature, which is exact for
polynomials of degree $\le 2M-1$.
Then, if $P_{h}(x)$ is such a polynomial, we have
\beq
  \f{1}{\Delta x}\int_{\bK}P_{h}(x)\,dx=\int_{I}P_{h}(\eta)\,d\eta=\sum_{q=1}^{M}w_{q}\,P_{h}(\eta_{q}).
\eeq

We let $M=N$ and $S_{N}=\hat{S}_{N}$, which is a
spectral-type nodal collocation DG approximation \citep{bassi:2013}, which is exact
for Cartesian coordinates and Legendre-Gauss quadrature.
Inserting Eq.~\eqref{eq:conservedNodalExpansion} into Eq.~\eqref{eq:semidiscreteDG},
letting $\phi(x)=\ell_{k}(x)$, using the quadratures defined above,
we obtain
\begin{align}
  \pd{}{t}\int_{\bK}\vect{U}_{h}\,\phi\,dx
  &=w_{k}\,\pd{}{t}\vect{U}_{k}\,\Delta x
  \label{eq:timeDerivativeTerm}
\end{align}
for the time derivative piece, where $\Delta x = x_{R} - \xL$.
Similarly, the last term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}
becomes
\beq
  \int_{K}\vect{F}(\vect{U}_{h})\,\pderiv{\phi}{x}\,dx
  = \sum_{q=1}^{N}w_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q}).
  \label{eq:volumeTerm}
\eeq
Now we may combine Eqs.~\eqref{eq:timeDerivativeTerm}-\eqref{eq:volumeTerm} in
Eq.~\eqref{eq:semidiscreteDG} resulting in the semi-discrete form
\beq
  \deriv{\vect{U}_{k}}{t}
   =-\f{1}{w_{k}\Delta x}
  \Big\{
  \Big[\,
    \widehat{\vect{F}}\ell_{k}\big|_{x_{R}}
     -\widehat{\vect{F}}\ell_{k}\big|_{\xL}
  \,\Big]
   -\sum_{q=1}^{N}w_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  \Big\}.
  \label{eq:semidiscreteDiscretized}
\eeq
We now have a system of ordinary
differential equations (ODEs), which may be evolved in time with an ODE solver.
In Section~\ref{sec:results} we use the third-order strong
stability-preserving Runge-Kutta (SSP-RK3) method \citep{shu:1988}.

\subsection{Slope Limiting}
\label{sec:Limiting}
A common feature of high-order numerical PDEs is unphysical oscillations in the
solutions around discontinuities.
It is therfore of great interest in the DG algorithm to implement slope limiting
of the polynomial $\mathbf{U}_h$. We use the total variation diminishing (TVD)
slope limiter \citep[see, e.g.,][]{cockburn:1998} in conjunction with the
troubled cell indicator discussed in \citet{fu:2017} to prevent excessive limiting
by flagging elements where limiting is needed. To do this, we need to reduce under- and
overshootings of the higher-order solution at cell boundaries compared to
the cell averages of neighbor cells. Recall from Eq.~\eqref{eq:conservedNodalExpansion}
that in each cell our solution is expressed in the nodal form. It is convenient,
however, for limiting purposes to express the solution in a modal representation:

\beq
\mathbf{U}_h (x,t) =
\sum_{l=1}^{N=k+1}c_{l}(t)\,P_{l}(x)
\eeq

\noindent where $P_{l}(x)$ are the Legendre polynomials. These
representations of the solution are equivalent by requiring weak equivalence

\beq
  \sum_{j=1}^{N}\int_{\bK} \left( \vect{U}_{j}(t)\, \ell_{j}(x) -
    c_{j}(t)\, P_{j}(x) \right) \phi\, dx = 0 \quad \forall\,\phi \in \mathbb{V}^{k}.
    \label{eq:NodalModal}
\eeq

\noindent Letting $\phi$ be the Lagrange interpolating polynomials $\ell_{i}$, we have

\beq
  \sum_{j=1}^{N}\int_{\bK}\ell_{i}(x)\, \ell_{j}(x)\, dx\, \vect{U}_{j}(t) =
  \sum_{j=1}^{N}\int_{\bK}\ell_{i}(x)\, P_{j}(x)\, dx\, c_{j}(t).
\eeq

\noindent Now letting $ M_{ij} = \int_{\bK}\ell_{i}(x)\, \ell_{j}(x)\, dx$ be the mass matrix,
$A_{ij} = \int_{\bK}\ell_{i}(x)\, P_{j}(x)\, dx$, $\bar{\vect{U}} = \{U_{1},...,U_{N}
\}$ be the nodal coefficients, and $\bar{\vect{c}} = \{c_{1},...,c_{N}\}$
be the modal coefficients, we see that the different representations are related by
a linear transformation

\beq
  M \bar{\vect{U}} = A \bar{\vect{c}}
\eeq
To perform slope limiting, we compare
the weight $c_{2}$, which is proportional to the first derivative of the
solution in the cell, to the neighboring cell averages by the following

\beq
  \mathcal{M}\,\widetilde{c}_{2} = \text{minmod}(\mathcal{M}c_{2},
    \beta_{\TVD}\mathcal{M}(c_{1}^{+} - c_{1}), \beta_{\TVD}\mathcal{M}(c_{1} - c_{1}^{-}))
  \label{eq:limiting}
\eeq

\noindent where $\widetilde{c}_{2}$ is the limited weight. The superscripts $-/+$
indicate the cell averages of the neighboring cells to the  immediate left/right
of the interface and $\mathcal{M}$ is a transformation matrix. For componentwise
limiting, we let $\mathcal{M}$ be the identity matrix. Notice that

\beq
%  \begin{align}
    \frac{1}{\Delta x}\int_{\bK}\vect{U}_{h}\,dx =
      \frac{1}{\Delta x} \sum_{l=1}^{N}\int_{\bK}P_{l}(x)\,\cdot 1\, dx\, c_{l}
      = \delta_{l1} c_{l} = c_{1}
  \label{eq:cellAverage}
%  \end{align}
\eeq
so $c_{1}$ is precisely the cell average. The minmod function in
Eq.~\eqref{eq:limiting} is defined as

@@

The parameter $\beta_{\TVD}$ takes values in the closed interval $\left[1,2\right]$
and scales the strength of the limiting. A minimal $\beta_{\TVD}$ corresponds with
the total variation diminishing scheme but is more
dissipative than the maximal $\beta_{\TVD}$ case, which is potentially more oscillatory.
Increasing $\beta_{\TVD}$ puts more weight on the neighboring cell averages,
making the minmod function more likely to set $\widetilde{c}_{2} = c_{2}$, when
no limiting is applied. If instead $\widetilde{c}_{2} \neq c_{2}$, the solution
is truncated to first order

\beq
  \vect{U}_{h} \to \widetilde{\vect{U}}_{h} = c_{1}\, P_{1}(x) + \widetilde{c}_{2}\,P_{2}(x)
\eeq
and, by Eq.~\eqref{eq:minmod}, $\widetilde{c}_{2}$ can potentially be zero as well.
Notice that, by Eq.~\eqref{eq:cellAverage}, the cell average is not altered by
the limiting process, so the process is conservatory.

In order to determine where slope limiting is necessary, we use the
troubled cell indicator (TCI) \citep{fu:2017} to prevent excessive limiting

@@

\noindent where $G\in\vect{G}\subseteq\vect{U}$ and $\vect{G} = \left(\rho,E\right)$.
Here, the sum in the numerator and the
max in the denominator are taken over the neighboring elements sharing a
boundary with target element $\bK$, $G_{\bK}$ is the cell average in $\bK$,
$G_{\bK}^{(j)}$ is the cell average
computed by extrapolating the polynomial representation from the neighboring
element $\bK^{(j)}$ into $\bK$, and $G_{\bK^{(j)}}^{(j)}$
is the cell average native to neighbor element $\bK^{(j)}$. An element is flagged
for limiting if, for any $G\in\vect{G}$, $I_{\bK}(G)>C_{\TCI}$, where
$C_{\TCI}$ is a defined threshold.

\subsection{Time Integration}
\label{sec:TimeInt}
Eq.~\eqref{eq:semidiscreteDiscretized} presented a system of ODEs that must
be evolved in time. We use the third-order strong stability-preserving
Runge-Kutta (SSP-RK) scheme \citep{shu:1988}.
The general $s$-stage Runge-Kutta time stepping algorithm, including
the limiting process, can be summarized as in \citet{cockburn:2001}:
\begin{itemize}
  \item[1.] Set $\bar{\vect{U}}^{(0)} = \bar{\vect{U}}^{n}$,
  \item[2.] For $i=1,\ldots,s$ compute:
  \begin{equation}
  \small  \bar{\vect{U}}^{(i)}
    = \Lambda^{\TVD}\Big\{\sum_{j=0}^{i-1}\alpha_{ij}\,\bar{\vect{U}}^{(j)}+\beta_{ij}\,\Delta t\,\bar{\vect{F}}\big(\bar{\vect{U}}^{(j)}\big)\Big\},
    \label{eq:rkStages}
  \end{equation}
  \item[3.] Set $\bar{\vect{U}}^{n+1}=\bar{\vect{U}}^{(s)}$.
\end{itemize}
Above, the TVD limiter limiter preventing unphysical
states is denoted by the operator $\Lambda^{\TVD}\{\}$. The SSP-RK3 coefficients $\alpha_{ij}$
and $\beta_{ij}$ may be found in Table~2.1 in \citet{cockburn:2001a}. For each step
in Eq.~\eqref{eq:rkStages}, the TVD limiter is applied to elements flagged
by the troubled cell indicator (where $I_{\bK} > C_{\TCI}$).
This algorithm is subject to a timestep stability condition $\Delta t_{\TVD}$
where $|\lambda|$ is the magnitude of the largest eigenvalue of the flux Jacobian.
Then (e.g., \cite{cockburn:2001a}):
\begin{equation}
  \Delta t_{\TVD} \le \f{1}{d}\,\f{(\Delta x/|\lambda|)}{(2k+1)}
  \label{eq:timestepRestrictions}
\end{equation}
where d is the dimension. For $d=1,k=2$ (third-order scheme) the effective Courant-Friedrichs-Lewy (CFL) factor is $0.2$.
In the numerical experiments presented in Section~\ref{sec:results} we use
a more conservative timestep restriction of $\Delta t = 0.1 (\Delta x/|\lambda|)$.

\subsection{Characteristic Decomposition}
\label{sec:characteristicDecomp}
Experience has shown that the slope limiting described in the previous section
is more effecient when performed on the so-called `characteristic variables`
as opposed to the conserved variables $\mathbf{U}_h$
\citep[see, e.g.,][for a description]{cockburn:1998}.
Inspired by theory for linear systems, we begin by
rewriting Eq.~\eqref{eq:conservation} in the quasi-linear form as follows
\beq
  \pderiv{\mathbf{U}}{t}
  + \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}}\pderiv{\mathbf{U}}{x}
  = 0.
  \label{eq:charEq}
\eeq
Because the Euler equations form a system of hyperblic
partial differential equations \citep[see, e.g.,][]{leveque:1992}, we can decompose the
Jacobian of the flux vector as

\beq
  \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}} =
  \mathcal{R} \Lambda \mathcal{R}^{-1},
\eeq
where the columns of $\mathcal{R}$ contain the right eigenvectors of the Jacobian,
the rows of $\mathcal{R}^{-1}$ contain the left eigenvectors, and
$\Lambda$ is a diagonal matrix containing the eigenvalues of the Jacobian.
For hyperbolic systems, the eigenvalues are real and the eigenvectors form a
complete set \citep[see e.g.,][]{leveque:1992}.
At this point, we will introduce the characteristic variable
$\mathbf{w} = \mathcal{R}^{-1}\mathbf{U}$. Multiplying both sides of Equation
\eqref{eq:charEq} by $\mathcal{R}^{-1}$ and simplifying, we are able to linearize the system of equations to
a system of advection equations
